---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(risoma)
```

# Introduction

The `risoma` package addresses the limitations of existing analytical tools in functional integration and parameter flexibility for microRNA isoform (isomiR) research.
The tool innovatively introduces a strategy of independently controlling insertion-deletion (InDel) and single nucleotide polymorphism (SNP) thresholds across distinct regions (5' end, central region, and 3' end) through an improved "seed-and-extension" algorithm, significantly enhancing the detection sensitivity for complex variants.
`risoma` integrates modules for miRBase database loading, isomiR identification, expression quantification (TPM normalization), and differential analysis.
It employs an S4 class system (*Isomir* and *IsomirDataSet*) for efficient cross-sample data management.
Compared to existing tools, `risoma` exhibits marked advantages in detecting central-region SNPs and parameter flexibility, covering more unique variant types.
`risoma` fills the gaps in functional comprehensiveness and user experience of current tools, providing robust technical support for unraveling isomiR regulatory networks.

# Data Structures

## Isomir Class

An instance of this class represents a microRNA isoform (isomiR) derived from a reference miRNA.
Its data structure includes the following key attributes:

-   **Reference Information**: Reference miRNA ID (*mature_id*), sequence (*mature_seq*), seed sequence(*seed_seq*)
-   **Alignment Features**: Template sequences containing flanking regions, sequencing read sequences and their abundance, insertion-deletion (InDel) counts at the 5' and 3' ends, mismatch (SNP) counts (including seed regions), sequence total distance, and a CIGAR (Compact Idiosyncratic Gapped Alignment Report) string detailing alignment specifics.
-   **Identifiers and Alignment Results**: Unique isoform ID (*ids*) and multiple sequence alignment (MSA) results against the reference miRNA.

## IsomirDataSet Class

This class integrates isomiR expression data across multiple samples.
Its core components include:

-   **Sample Metadata** (*sample_info*): A data frame storing sample names, experimental groups, and file names.
-   **Group-level isoforms** (*group2isoforms*): A list of isoforms grouped by experimental groups (e.g., tissue types or treatment conditions).\
-   **Isoform Expression Matrices**:
    -   *Sample-level matrix* (*sample_expr*): Rows represent isoforms, columns represent samples, with values indicating transcripts per million (TPM) normalized expression levels.
    -   *Group-level matrix* (*group_expr*): Aggregated TPM mean values based on experimental groups (e.g., tissue types or treatment conditions).
-   **Isoform Clustering Information** (*ref2isomir*): A list of isoforms grouped by reference miRNA to support cluster analysis based on sequence similarity.

# Construction of isomiR Expression Profiles

## Data input

```{r}
sample_info_file <- system.file("extdata", "sample_info.csv", package = "risoma")
fq_dir <- system.file("extdata", "fastqs", package = "risoma")
```

### Input File Specifications

-   **Sample Metadata File (sample_info_file)**: The metadata file must include the following columns:

    -   Column 1: Sample names.
    -   Column 2: Filenames of sequencing data (FASTQ or compressed FASTQ files).
    -   Column 3: experimental groups.

-   **Data Directory**: Sequencing data files (FASTQ or compressed formats) must be stored in a user-specified directory.

-   **Reference miRNA Database**: Species-specific miRNA information is retrieved from miRBase using the `load_mirnas(species_code, seed_len)` function.

    Parameters:

    -   `species_code`: miRBase species identifier (e.g., "ath" for *Arabidopsis thaliana*).
    -   `seed_len`: Seed sequence length (default: 16 nucleotides).

## Load microRNA data from miRBase

```{r load miRNAs}
vvi_mirnas <- load_mirnas("vvi") 
head(vvi_mirnas)
```

This function retrieves microRNA (miRNA) data from the miRBase database for a specified species.
It extracts mature miRNA sequences, seed sequences, and flanking regions, and returns the data in a structured data frame.

Reference miRNA sequences for grape were loaded from miRBase via the `load_mirnas("vvi")` function, yielding 186 mature miRNA sequences and their precursor information, including seed sequences and flanking regions.

## Detect isomiRs

```{r detect isomiRs}
isomir_dataset <- detect_isomirs(sample_info_file, fq_dir, vvi_mirnas, min_tpm = 5)
```

This function detects isomiRs from small RNA-Seq data for multiple samples.
It processes FASTQ files, identifies isomiRs, and generates expression profiles for samples and groups.
The results are returned as an IsomirDataSet object, which can be used for further analysis.

The `detect_isomirs()` function was executed with the following parameters: expression threshold ≥5 TPM (transcripts per million), maximum InDels of 3 at the 5' and 3' ends, and a maximum of 1 SNP per region (5'/3' ends and central seed region).

The resulting *IsomirDataSet* object integrated sample-level and group-level expression matrices, alignment details, and metadata.

## Count isoforms per reference miRNA

Further analyses are performed on an IsomirDataSet object constructed using publicly available small RNA sequencing datasets from the Gene Expression Omnibus (GEO) database (accession: GSE59802).
This dataset comprises 50 samples across 25 critical developmental stages, with two biological replicates per stage.

```{r}
data("isomirs")
```

```{r get isoform number}
get_isoform_num(isomirs)
```

This method calculates the number of isoforms associated with each reference miRNA in an IsomirDataSet object.
It returns a vector of counts for reference miRNAs.

## Calculate the number of expressed isoforms per group

```{r calc expression number}
expr_num <- calc_group_expr_num(isomirs)

head(expr_num)
```

This method calculates the number of expressed isomiRs for each group in an IsomirDataSet object.
It categorizes isomiRs into three types:

-   Total: The total number of isomiRs expressed in each group.

-   Organ-Specific: The number of isomiRs expressed in only one group.

-   Common: The number of isomiRs expressed in all groups.

Plot total number of expressed isoforms per group:

```{r plot total number of isoforms, fig.width=5, fig.height=3}
library(ggplot2)

expr_num <- calc_group_expr_num(isomirs)
ggplot(expr_num, aes(x = group, y = total)) + 
  geom_bar(stat = "identity") + 
  coord_flip() +
  xlab("") + ylab("number of isomiRs") + 
  labs(title = "number of isomiRs at each developmental stage") +
  theme(text = element_text(size = 12), plot.title = element_text(hjust = 0.5))
```

Plot the number of isoforms that are specifically expressed in each experimental group:

```{r plot the number of specifically expressed isoforms, fig.width=5, fig.height=3}
ggplot(expr_num, aes(x = group, y = organ_specific)) + 
  geom_bar(stat = "identity") + 
  coord_flip() +
  xlab("") + ylab("number of isomiRs") + 
  labs(title = "Distribution of organ-specific isomiRs") +
  theme(text = element_text(size = 12), plot.title = element_text(hjust = 0.5))
```

## Tissue Specificity Index (TSI)

To quantitatively assess the spatiotemporal specificity of isomiR expression across tissues or experimental conditions, `risoma`employs the Tissue Specificity Index (TSI), a metric that evaluates the bias in expression distribution.
The TSI ranges from 0 to 1, where:

-   TSI = 0: Indicates uniform expression across all tissues/conditions.

-   TSI = 1: Indicates exclusive expression in a single tissue/condition.

```{r}
tsi_df <- calc_tsi(isomirs)

head(tsi_df)
```

Generates a violin plot with overlaid boxplots to visualize the distribution of TSI values for reference miRNAs and isoforms.

```{r plot TSI, fig.width=5, fig.height=3}
ggplot(tsi_df, aes(x = type, y = tsi)) + geom_violin() + 
  geom_boxplot(width = 0.05, fill = "black", outlier.color = NA) + 
  stat_summary(fun = median, geom = "point", fill = "white", shape = 21, size = 2.5) +
  xlab("isomiR type") + ylab("TSI") + 
  theme(text = element_text(size = 12))
```

## Inspect a specific isomiR

Extract isomiR data for a specific reference miRNA.
For example `vvi-miR156f` and its isoforms were analyzed by:

```{r vvi-miR156f}
isomir_156<- get_isomir_by_ref(isomirs, "vvi-miR156f")

isomir_156
```

Expression data were exported via:

```{r miR156f expr}
expr_156 <- get_expr_by_ref(isomirs, "vvi-miR156f")
expr_156
```

This method extracts expression data for isomiRs associated with a specified reference microRNA.
It is useful for analyzing the expression patterns of isomiRs derived from a specific microRNA across different conditions or groups.

Plot heatmap of expression values for isoforms of `vvi-miR156f`:

```{r, fig.width=5, fig.height=3}
pheatmap::pheatmap(expr_156, scale="row", fontsize_row = 10, width=6, height=5)
```

Extract aligned isoforms to a `vvi-miR156f`:

```{r}
get_alignment_by_ref(isomirs, "vvi-miR156f")
```

This method extract the alignment of the sequences of isomiRs and their reference template sequence using multiple sequence alignment (MSA).
It is useful for visualizing sequence variations and identifying potential editing events in isomiRs.

# Perform differential expression analysis on isomiRs

```{r}
data("isomirs")
```

`get_deg_data` extracts expression data and sample information from an `IsomirDataSet` object for specified treatment and control groups, preparing it for differential expression analysis.

```{r get deg data}
deg_data <- get_deg_data(isomirs, treatment = "Flower_FB", control = "Inflorescence_WD")
```

```{r}
dds <- DESeq2::DESeqDataSetFromMatrix(countData = deg_data$expr,
                                      colData = deg_data$sample_info,
                                      design = ~ group)
dds <- DESeq2::DESeq(dds)
res <- DESeq2::lfcShrink(
  dds,
  coef = DESeq2::resultsNames(dds)[[2]],
  type = "apeglm",
  quiet = TRUE
)
res <- res[!(res$baseMean == 0 &
              is.na(res$padj) & is.na(res$pvalue) &
              is.na(res$log2FoldChange)), ]
res <- res[!res$log2FoldChange == 0, ]
res <- res[!(is.na(res$pvalue) & is.na(res$padj)), ]
res <- res[!is.na(res$padj), ]
res <- res[res$padj <= 0.05, ]

result <- as.data.frame(res[order(res$log2FoldChange), ])
head(result)
```
